---
title: "Pollution, plotting and packages"
output: html_notebook
---

In this notebook we cover:

* How to install R *packages*.
* Making plots with regression lines.
* Dealing with dates and times.
* (Optional) Averaging over time periods.

The last part of this notebook is optional, starting at "Getting fancy with
dates and times".  Please feel free to have a go at those final sections if you
have time and / or interest.  We won't assess your "best effort" on that last
part of the notebook.

## London pollution dataset

Our example dataset records various air quality measures from the Marylebone
Road, in London:

```{r}
# Run this chunk.
# "london_df" is the London dataset data frame.
london_df <- read.csv('air_pollution_hourly_98-05_london.csv')
# Show the first six rows.
head(london_df)
```

The first six rows show you that the dates start in 1998.  Now show the *last* six rows to see when the dates end:

```{r}
#- Show the last six rows of london_df
...
```

You should find that these last rows have dates from 2005.

Check the structure of the `london_df` data frame using the `str` function:

```{r}
str(...)
```

The data frame has the following columns:

* `date` : the date of the recording, as a UK formatted date string.
* `ws`: wind speed.
* `wd`: wind direction in degrees.
* `nox`: nitrogen oxides.
* `no2`: nitrogen dioxide.
* `o3`: ozone.
* `pm10`: 2.5 - 10 micrometer particulate level.
* `so2`: sulphur dioxide.
* `co`: carbon monoxide.
* `pm25`: >2.5 micrometer particulate level.

We will come onto the date soon.

## Packages

R has a huge number of *packages*.  Packages are collections of code
that integrate with R, and so give you more functions to work with many
forms of data.

In your R life, you will find yourself using packages very often.  Among the
things that packages can do for you, are reading data in almost any format you
can imagine, adding powerful tools for plotting, filtering, and rearranging
data, and implementing a vast array of statistical tests.

For this tutorial, you need two packages to make the best use of the `date`
column in our data frame.  These are:

* The [lubridate](https://lubridate.tidyverse.org) package to "Make
  Dealing with Dates a Little Easier".  Lubridate has very useful
  functions for processing dates, as you will soon see below.
* The [xts](https://cran.r-project.org/web/packages/xts) package.  XTS stands
  for eXtensible Time Series. The `xts` package provides many useful functions
  for dealing with data like ours, that have dates and times.

There are two typical steps to using a package in R.

1. *Install* the package using the `install.packages` function.  You need to do
   this only once for a particular installation of R.  This means that, until
   you install another, new copy of R on your computer, you don't need to run
   `install.packages` again for a package you have already installed. If you
   want to use a package, you always have to install it.
2. *Use* the package using `library(my_package_name)`.  You do this once in
   each notebook or R session, before you use the functions from the package.
   This is the typical way of using a package, but there are other ways.

Let's start by doing that routine for the Lubridate package.

First you run `install.packages` using something like the chunk below.

**Don't run the chunk yet**.

Notice that the package name goes in between quotes -- 'lubridate' --- to tell
R that this is a piece of text --- in this case, the name of the package we
want R to fetch from the web and install on our computer.

Now run the chunk below:

```{r}
install.packages('lubridate')
```

You should see various bits of output as R goes to get the package from the
web, then builds it into something it can use on this computer.

Much of this output is red text; this is normal. But, **Call us over if you get
an error**.

After this is finished you should be able to run the following chunk.  This
makes the Lubridate package available to the rest of the R session in the
notebook.

```{r}
library(lubridate)
```

**Ignore any warnings** that you get about `masked` objects, they are normal.

**Call us for help if you get an error**.

Notice that in the code above, `lubridate` is *not* between quotes. This is
because, after you installed the package, R knows that the name `lubridate`
refers to the package.

Now make sure that all is working by trying a function from the Lubridate
package.  One of the functions in the Lubridate package is
[parse_date_time](https://lubridate.tidyverse.org/reference/parse_date_time.html).

Run the following chunk that uses `parse_date_time` to convert some text into
an R date value (more on this later):

```{r}
a_date <- parse_date_time('20-02-14 14:00', 'ymd-HM')
a_date
```

**Call us for help if you get an error**.

Now your turn.  Later in this notebook, you will also need the `xts` package.

Remember, there are two steps:

1. *Install* (once for each installation of the R language).
2. *Use* (once for each notebook / R sessions).

Look at the chunks above for inspiration, where we installed and then used the `lubridate` package.

First the install:

```{r}
#- Use install.packages to install the 'xts' package.
...
```

Next use the package:

```{r}
#- Use the library function to use the xts package in this R notebook
#- session.
...
```

Ignore any warnings that you get about `masked` objects, they are normal.

Finally, the chunk below tests that everything went OK with your `xts`
installation, and your `library` function call above.  The chunk use the
`is.xts` function from the `xts` package to tell you whether `a_date` (see
above) is an eXtensible Time Series (it isn't).  Run the chunk to confirm that
`xts` is properly installed.  It should show FALSE.

```{r}
# Run this chunk.
is.xts(a_date)
```

## A note for Windows users

Windows makes things a little harder to install packages.  For some packages -
**not the ones above** - you will need a system called RTools.  See [this
RTools page](https://cran.r-project.org/bin/windows/Rtools) for instructions
on installing RTools.

**You do not need to install RTools for this notebook**, but you may want to
install it when you try installing R packages other than the ones here -
packages other than `lubridate` and `xts`.

## Labels for plots

Returning to the dataset, we might first want to look at the histogram of `nox`
values:

```{r}
#- Show a histogram of the "nox" column of "london_df".
hist(...)
```

We are also going to be interested in the `pm25` values from this data frame.

```{r}
#- Show a histogram of the "pm25" values from the "london_df" data frame.
...
```

These histograms are OK, for a default, but when we get to presenting data to
someone else, we will want to make the labels more meaningful. First we might
want to set the title.  We use the `main=` option to the plotting command, like
this:

```{r}
hist(london_df$nox, main='Marylebone Road NOx levels')
```

The x-axis label is also a bit hard to read.  Tadaa! - we use the `xlab=`
option to the plot command:

```{r}
hist(london_df$nox, main='Marylebone Road NOx levels', xlab='NOx level')
```

Show a histogram of the `pm25` column from `london_df`, with a suitable title
(using `main=`) and x axis label (using `xlab=`)

```{r}
hist(...)
```

Sometimes these plot commands can get long, with many arguments.

R allows us to spread a function call across many lines.  It knows all the
lines are part of one function call, because of the opening `(` and closing `)`
parentheses around the *arguments* to the function. The arguments are the
things that go between the parentheses, separated by commas, to tell the
function what to do.

For example, I could spread the previous histogram function call over three
lines, like this:

```{r}
# Spreading a function call over three lines.
# The parentheses allow R to see that all three lines are for a single
# function call.
hist(london_df$nox,
     main='Marylebone Road NOx levels',
     xlab='NOx level')
```

Spread your histogram call for the `pm25` values (above) over three lines:

```{r}
#- Make a histogram of the "pm25" values with a title and x-axis label.
#- Spread the function call over three lines.
#- Don't forget the commas after each argument, between the brackets.
hist(...
     ...
     ...)
```

## The same arguments work for many plot types

We have so far been doing histograms, but the same options apply to scatter
plots:

```{r}
# Plot of no2 against nox.
# "main" and "xlab" arguments to a scatter plot.
plot(london_df$nox, london_df$no2,
     main='Marylebone Road NOx levels',
     xlab='NOx level')
```

Now we have an ugly y-axis label.  You can guess what we need to do.

```{r}
# "main" and "xlab" and "ylab" arguments to a scatter plot.
plot(london_df$nox, london_df$no2,
     main='Marylebone Road NO2 against NOx levels',
     xlab='NOx level',
     ylab="NO2 level")
```

Now use the `main` and `xlab` and `ylab` options to plot the `nox` values
against the `pm25`, with suitable title, x and y axis labels.

```{r}
#- Plot nox on the x-axis and pm25 on the y axis.
#- Use "main" and "xlab" and "ylab" arguments to a scatter plot.
#- You might want to split your function call over several lines.
plot(...)
```

## Drawing a regression line

The plot of "nox" and "pm25" suggests there might be some sort of straight-line
relationship between the two.

We haven't covered how to find a "good" line for two sets of points, and there
are different ways on deciding on a good line.  The most common way is called
the *least squares* regression line.  This often appears on plots labeled
humbly as a "trend line".

We can define any line on a graph with an *intercept* and a *slope*.

* The *intercept* is the value on the y axis where the line crosses the y axis.
  It is therefore the value for the y axis when the corresponding x value is
  zero.
* The *slope* is the amount the y value goes up for every one unit increase in
  x. The slope can be negative, in which case the y value goes down for every
  one unit increase in x.

The line, defined by the intercept and slope, gives a prediction for each
y value, given the corresponding x value.

This next paragraph is for information --- you will hear more about this in
your statistics lectures.  The "best" line is the line that gives the best
prediction of the y values, in some sense of "best". One common measure of
"best" is called *least squares*.  The *least squares* line is the "best" line
in the sense that it is the line that gives the smallest squared distance of
the actual y values from the predicted y values given by the line.

R was born to solve problems like this.  Here we calculate the *least squares*
line using R's `lm` function.  `lm` stands for Linear Model.  In this case
R works out the best *least squares* intercept and slope for the `nox` column
data and the `pm25` column data.

```{r}
best_line <- lm(london_df$pm25 ~ london_df$nox)
best_line
```

Notice the tilde `~`, and the order of the columns.  The `~` means "as
a function of" so `lm(london_df$pm25 ~ london_df$nox)` asks R to find a Linear
Model (straight line) that *predicts* the `pm25` values *as a function of* the
`nox` values.  That is, it is a line predicting the y values (`pm25`), on the
**left** as a function of the x values on the **right**.

Notice too from the output that it has worked out the intercept and the slope.

We can put this linear model line on the plot, using a function called `abline`
--- like this:

```{r}
plot(london_df$nox, london_df$pm25)
abline(best_line)
```

Now your turn.  First get the best line for the `pm10` values *as a function
of* the `nox` values.  Remember which goes on the left and which on the right.

```{r}
best_pm10_line <- ...
# Show the calculated line parameters.
best_pm10_line
```

Now do a plot with the `nox` values on the x axis, and the `pm10` values on the
y axis.  Add a title and labels for the two axes.   Add the line you have just
calculated.

```{r}
#- Plot nox on the x-axis and pm10 on the y axis.
#- Add title, x-axis and y-axis labels.
#- Add the least squares line.
plot(...)
...
```

## Dates and times

We often need to deal with datasets where the observations have dates and
times.

This pollution dataset is a typical example.

```{r}
# Show the first six rows of the pollution dataset.
head(london_df)
```

The first column in this dataset is `date`.  In fact we can see that it is text
that contains both a date and a time in hours:minutes.

As usual, when R reads in the data with `read.csv`, it finds that the first
column looked like text (characters), and makes it into a `factor`.  We can see
this is the case with `str`:

```{r}
str(london_df)
```

Show the *first six* levels of the factor contained in the `date` column of
`london_df`.

**Hint** : You know how to show the levels of a factor.  Remember how we show
the first six of anything?

```{r}
head(...)
```

The factor in `london_df$date` is a pretty useless factor, because each
observation is one level. This must be so, because all the date / time values
are unique.  Here we use the `nrow` function to show the number of rows in the
data frame.

```{r}
nrow(london_df)
```

Notice that this is also the first value we get back from the `dim` function,
that you may remember from the tutorials:

```{r}
dim(london_df)
```

You can see that the number of rows in the data frame is the same as the number
of levels in the factor, listed from `str` above.

One problem with this huge factor, is that it is very inconvenient to do plots
against date and time.  You might think of trying it now, but R will get stuck
for a while, while it tries to deal with the factor having many thousands of
levels.

We would like to make `date` into something more useful than this enormous
factor.

It is probably not surprising to you, that R has functions for converting
between factors (which is what we have now) and character strings, and
functions for converting character strings to something that can represent
a date and time.

To start, let's first convert the nasty factor back into character strings, as
the values were in the original file.

To do this, I force the values to be characters, with the R `as.character`
function, and put them back into the data frame column:

```{r}
# Set the date column to be bits of text (character strings).
london_df$date <- as.character(london_df$date)
head(london_df$date)
```

Show what type of values `date` contains now, using `str`:

```{r}
...
```

We want to convert these character strings into something R understands as
a date and a time.  You may remember we used the Lubridate function
`parse_date_time` for this.  Here is the example from further up the notebook.


```{r}
a_date <- parse_date_time('20-02-14 14:00', 'ymd-HM')
a_date
```

If you look at the [documentation for
parse_date_time](https://lubridate.tidyverse.org/reference/parse_date_time.html),
you will see that we need to pass at least two things to the `parse_date_time`
function. The first is the bit of text, or vector of bits of text, that we
want to convert into dates.  This is called `x` in the documentation.  The
second is something called `orders`.   It specifies the order in which the
various parts of the date and time occur in the string.

Here is another version of the example above.  The only difference is that we
first put the text into a variable `my_date_str` and then convert that text
into an R date.

```{r}
# Run this chunk.  Don't change it.
my_date_str <- '20-02-14 14:00'
a_date <- parse_date_time(my_date_str, 'ymd-HM')
# Show the resulting R date.
a_date
```

In this example, we said that the text "20-02-14 14:00" (from the variable
`my_date_str`) had *order* "ymd-HM". This means that the first thing in the
text is the Year ('y'), the second thing is the Month ('m') and the third
thing is the Day ('d'). After that we have the Hour ("H") and the Minute
("M").

But now - what if I meant to say that "20-02-14" actually meant the 20th of
February 2014 (but I still meant the time to be 14:00).  How would I change
"order" to do that?

```{r}
#- Change the "order" argument below to tell parse_date_time
#- that 20 is the day of the month, and 14 is the year.
#- If you get an error "object 'my_date_str' not found", make
#- sure you have run the chunk above this one.
another_date <- parse_date_time(my_date_str, 'ymd-HM')
# Show the new date.
# If you got this right, you should see the year is 2014.
another_date
```

These are the codes for parts of the date and time that you have seen:

* `d` meaning two-digit number for day of the month.
* `m` as two-digit number of month of the year.
* `y` for two-digit number of year of the century.
* `H` for two-digit hour.
* `M` for two-digit minute.

You can see all the character codes in the documentation for the R function
[strptime](https://rdrr.io/r/base/strptime.html).  Notice the codes can have
`%` in front of them, if you like, and as they do in that documentation. So,
the following chunk is the same as the example you have already seen:

```{r}
# With the order codes having percents in front - same thing as without.
a_date_again <- parse_date_time('20-02-14 14:00', '%y%m%d-%H%M')
a_date_again
```

Notice too that the result of `parse_date_time` is a new type of thing,
`POSIXct POSIXt`, where the final 't' stands for Time.  This is one of R's ways
of storing a date and time.

```{r}
class(a_date)
```

As you will soon see, this type of thing is much more useful to us, because we
can do fancy things like plot values over time.

## Using dates and times

We have found that we could convert the dates in the given format with
`parse_date_time`.

We want to convert the date characters in `london_df$date` into proper R dates.

To start with, let's play with an example string, by getting the first value in
the `london_df` `date` column:

```{r}
first_date_string <- london_df$date[1]
first_date_string
```

We confirm this is text (characters) with `class`:

```{r}
class(first_date_string)
```

Here the first number refers to the day of the month, the second to the month
of the year, and the third to the year, followed by the hour and minutes.

Use `parse_date_time` and a suitable `order` to convert `first_date_time` to an
R date:

```{r}
first_date <- parse_date_time(first_date_string, ...)
# Show the date.
first_date
```

You should see that the year for this date is 1998.

In fact, we can do this all in one shot, using R's idea about using *vectors*.
Fill in the correct value for `order` and run this chunk to convert all the
bits of text in the `london_df$date` column into proper R dates:

```{r}
london_df$date <- parse_date_time(london_df$date, ...)
# Show the first six rows.
head(london_df)
```

We can now use these dates for plotting.

Plot the dates in the `date` column of `london_df` on the x axis with the
values from the `pm10` column on the y axis.  Add a title and an x axis label
and a y axis label.

```{r}
plot(...)
```

## Getting fancy with dates and times

**This part of the notebook, and everything following, is optional for your
assessment**.

The trick for using dates to their best potential is to make our data into
a *timeseries*.

In particular, we are going to use the specialized `xts` R *package* to make
useful timeseries.

An Xts time series consists of two things:

* A vector of dates or date / times, called the *index*
* A *matrix* of data where each row corresponds to the matching row in the
  index.

A *matrix* in R is like a data frame, in that it has rows and columns, but
unlike a data frame, all the values have to be of the same type (such as number
or string or date).

That's all you need about matrices for now. In this tutorial you can follow the
steps here to make an Xts time series.

First we need the number of columns in the data frame.  We can get this using
`dim`, like this:

```{r}
dim(london_df)[2]
```

This gives us the second value returned from `dim`, which is the number of
columns.  We can also use the `ncol` function, like this:

```{r}
ncol(london_df)
```

We convert the last 9 columns in the data frame into a *matrix*.

```{r}
# A matrix of data. This has all rows, and all columns except the first.
# Number of columns.
n_cols <- ncol(london_df)
# Data frame without the first column.
df_without_first_col <- london_df[,2:n_cols]
# Data frame converted to matrix
data <- as.matrix(df_without_first_col)
head(data)
```

Show the type (class) of this new thing `data`:

```{r}
...
```

We had to drop the first column of `london_df`, because the first column
contains dates.  Remember that matrices must contain data of the same type.  If
we had left in the first column, before converting to a matrix, R makes
everything into a character string, because that's the only type that works for
all the dates and the numbers.

```{r}
# Convert the original data frame, including the first column.
ugly_matrix <- as.matrix(london_df)
head(ugly_matrix)
```

Now we can make the Xts object from the *index* and the *matrix*:

```{r}
# The Xts object.
london_ts <- xts(data, london_df$date)
head(london_ts)
```

Don't worry about the warning about the timezone being different from the
current timezone.

The extended timeseries allows us to make even nicer plots, where the plots
know about the dates, and where the dates go on the x-axis by default.

```{r}
plot(london_ts$nox)
```

```{r}
plot(london_ts$o3)
```

But we can also *aggregate* over days and months and years.  Xts has some nice
functions for this:

```{r}
by_month <- apply.monthly(london_ts, mean, na.rm=TRUE)
head(by_month)
```

```{r}
plot(by_month$nox)
```

```{r}
by_year <- apply.yearly(london_ts, mean, na.rm=TRUE)
head(by_year)
```

```{r}
plot(by_year$nox)
```

```{r}
plot(by_year$o3)
```

## Your turn

You will need to guess or web-search or read the help on xts to work out what
function you need to aggregate by day.

```{r}
#- Aggregate "london_ts" by day, to get the mean pollution level
#- on each day.
by_day <- ...
# Show the first six observations
head(by_day)
```

```{r}
#- Plot the PM10 levels from the daily aggregation above.
plot(...)
```

## The end

That's it.  **You don't need to submit this notebook**.  But - for practice,
please consider doing the submission checks that you will also do for your
homework.  Specifically: make sure that all the chunks execute correctly.  To
do that, go to the Run button at the top of the notebook, and click "Restart R
and Clear Output".  Then go to the top of the notebook, and run each chunk in
turn, to make sure that they run without error, and give you the output you
expect.  *Remember, for your homework, we going to run an automated marking
script over this file, so it has to execute correctly, for you to get the
marks*.

But **please don't submit this notebook, it's just for practice**.